---
title: "Human Actvitiy Recognition"
author: "Joben Ilagan"
date: "19 October 2015"
output: html_document
---

Six young health participants were asked to perform one set of 10 repetitions of the Unilateral Dumbbell Biceps Curl in five different fashions: 

* exactly according to the specification (Class A), 
* throwing the elbows to the front (Class B), 
* lifting the dumbbell only halfway (Class C), 
* lowering the dumbbell only halfway (Class D) and 
* throwing the hips to the front (Class E).

They measured their movements and the data can be found here: 
http://groupware.les.inf.puc-rio.br/har.

Read more: http://groupware.les.inf.puc-rio.br/har#sbia_paper_section#ixzz3oyROG8S5

The goal of this project is to predict the manner in which they did the exercises, that is, whether correctly (Class A) or incorrectly (Classes B to E). This is the "classe" variable in the training set. 

The other variables may be used for the prediction.

```{r}
library(caret)

training<-read.csv("pml-training.csv")
testing<-read.csv("pml-testing.csv")
pctTraining<-dim(training)[1] / (dim(training)[1] + dim(testing)[1])
pctTesting<-dim(testing)[1] / (dim(training)[1] + dim(testing)[1])
```

## Exploratory Data Analysis
The training set has `r dim(training)[1]` rows and `r dim(training)[2]` columns. The training set has `r pctTraining*100`% of the total data.

The testing set has `r dim(testing)[1]` rows and `r dim(testing)[2]` columns. The testing set has `r pctTesting*100`% of the total data.

We can therefore extract more data from the training set for cross-validation and (internal) testing before using the actual testing data set (pml-testing.csv) at the end of the exercise. 

The variable *classe* has five values as shown here:

```{r}
levels(training$classe)
```

## Expectations

A cross-validation sample is taken from 20% the training set to estimate the out-of-sample error.

Likewise, a testing sample (apart from the actual test data) is taken from another 20% of the training set.

```{r}
inTraining<-createDataPartition(training$classe,p=0.6,list=FALSE)
trainingML<-training[inTraining,]
temp<-training[-inTraining,]
inTesting<-createDataPartition(temp$classe,p=0.5,list=FALSE)
testingML<-temp[inTesting,]
xvalidationML<-temp[-inTesting,]
```

Training, Cross-Validation and Testing should now have a 60-20-20 mix.

```{r}
pctTrainingML<-dim(trainingML)[1] / (dim(trainingML)[1] + dim(testingML)[1] + dim(xvalidationML)[1])
pctTestingML<-dim(testingML)[1] / (dim(trainingML)[1] + dim(testingML)[1] + dim(xvalidationML)[1])
pctXValidationML<-dim(xvalidationML)[1] / (dim(trainingML)[1] + dim(testingML)[1] + dim(xvalidationML)[1])
pctTrainingML
pctTestingML
pctXValidationML
```
